Building an SSI Cluter with GNU/Hurd

### 2019-05-19 libpager/paper.tex
a checkpoint of work on "Building an SSI Cluster" paper

\usepackage{nopageno}
\usepackage{vmargin}
\usepackage{comment}

\usepackage{../slides/hurd}
\usepackage{tikz}
\usetikzlibrary{positioning, fit, backgrounds, arrows, trees, shapes}

@@

\title{Building an SSI Cluter with GNU/Hurd}
\author{Brent Baccala}
\date{}

@@

% Shortcuts just for this document
\def\libpager{{\tt libpager\ }}

@@

Operating system designers have tried for years to leverage
the power of multiple computers and develop cluster
operating systems.  Notable attempts include VAX/VMS, various cluster
Linux systems [MOSIX, refs], Beowulf, Plan 9, ScaleMP, and
Hadoop.

@@

GNU/Hurd is an operating system based on the Mach kernel, which was
designed for use in a cluster environment, making it a natural choice
to build a POSIX-compatible SSI cluster.  Development on Hurd has
progressed slowly, but a Debian port has been available since 2013,
and approximately three quarters of the Debian packages compile and
run on Hurd.

To support a multi-node Hurd cluster, the libpager used by filesystem
translators needs to support multiple clients (i.e, multiple kernels
mmap'ing files).  Hurd's original libpager, written in the mid-1990s,
only supported a single client, and was thus suitable for single node
operation.

This is a crucial library.  It provides the cache coherence logic
necessary to implement inter-node shared memory, whether disk-based or
not.  Its use in the tmpfs translator, which is in turn used by
Hurd's POSIX shared memory implementation

Hurd's netmsg and libpaper have been developed to the point where Hurd
could be usable as a cluster operating system, though the current
Debian distribution doesn't use this functionality.  At the time of
this writing (2019), the major barriers to building a Hurd cluster are
the 32-bit address space and the lack of SMP support, enhancements
that are beyond the scope of the work described in this paper.

The CAP theorem states that a distributed system can't have all of
three things: consistency, availability, partitioning.  Right now, we
sacrifice availability -- if the network partitions, libpager can
deadlock.  If we introduce a timeout on lock requests, then we're
sacrificing consistency.

@@

Mach's ability to attach out-of-band memory to messages is thus a
crucial component in the efficient implementation of its memory
management.  A page is typically supplied to kernel by attaching it to
a memory_object_data_supply message.  No copying needs be done, as the
page is transferred to kernel control simply by adjusting bits in the
hardware memory management tables.  Likewise, transferring a modified
page from the kernel to a user-space memory manager is done by
attaching it to a Mach message in a similar way.

@@

@@

\section{The NETMSG Server}

To implement distributed virtual memory in Hurd, we need to transport
Mach messages across the network.  Several programs have been written
over the years to transport Mach messages over a network connection;
all collectively go under the name of the ``netmsg server'', or just
``netmsg''.

I wrote Hurd's current netmsg server in 2016.  It operates over a
single unsecured TCP/IP connection and has many limitations, but it
works well enough to implement shared virtual memory.  In server mode,
it presents a single node, by default the root of the filesystem, to
any clients that connect.  In client mode, it acts as a translator,
allowing a remote server's exported node to be mounted in the
local filesystem.

Out-of-band memory is simply
copied across the network connection, then deallocated.

\section{LIBPAGER}

Hurd's implementation of the Mach memory management protocol is
implemented in a library called ``libpager''.  The original libpager,
written in the mid-1990s, only supported a single client (i.e, a
single Mach kernel), and was thus suitable for single node operation.
Its API, however, is flexible enough to allow a more sophisticated,
multi-client libpager as a drop-in replacement.  For dynamically
linked servers, all that is required is to update the system shared
library to obtain multi-node functionality.

@@

This routine is expected to allocate memory (unless it returns an
error) containing the requested data and pass it to \libpager, which
is responsible for deallocating it (typically by passing it to the
kernel).  A common use scenario is that {\tt pager_read_page()}
reads a page from disk and returns it.  {\tt pager_read_page()}
can be called from multiple threads to service multiple paging
requests, and must implement its own thread safety.

@@

This routine is expected to save a modified memory page just before
\libpager releases its last copy of the data.  It does no memory
allocation or deallocation.  It receives allocated memory and leaves
it allocated, with \libpager responsible for dellocation, or passing
it to the kernel, if a new request has arrived in the time required
for {\tt pager_write_page()} to complete

@@

How do we figure out which client is the local kernel?  One possible
solution is for netmsg to listen on /servers/netmsg for an RPC that
accepts a send right and identifies which node currently hosts the
corresponding receive right.

@@

DISTRIBUTED LIBPAGER.  We could use a shim process on remote nodes, to
allow requests for an existing page go to a node that holds a copy of
that page, instead of to the node with the disk.  Two ways I can
imagine this.  Either ext2fs forks off multiple processes on different
nodes, or a mechanism is developed to detect when a port is remote,
figure out which node it (currently) resides on, and then libpager can
fork a process and push it to the other node, invisible to ext2fs
proper.

\section{Configuration Options}

Several implementation choices exist, and perhaps should be configuration options.

@@

@@

@@

\section{The Pagemap}

All Mach memory management operations operate on memory pages, so
\libpager needs to maintain a table that tracks the state of each
page, called the pagemap.  Hurd's original \libpager had a fairly
simple pagemap.  The pagemap was an array of 16-bit shorts; each page
had a single 16-bit entry, allocated as follows:

\begin{verbatim}
#define PM_WRITEWAIT  0x0200   /* queue wakeup once write is done */
#define PM_INIT       0x0100   /* data has been written */
#define PM_INCORE     0x0080   /* kernel might have a copy */
#define PM_PAGINGOUT  0x0040   /* being written to disk */
#define PM_PAGEINWAIT 0x0020   /* provide data back when write done */
#define PM_INVALID    0x0010   /* data on disk is irrevocably wrong */

@@

enum page_errors
{
  PAGE_NOERR,
  PAGE_ENOSPC,
  PAGE_EIO,
  PAGE_EDQUOT,
};
\end{verbatim}

Four bits are allocated for error code.  (all other errors get collapsed into EIO)

The remaining bits indicate if the page is in core, if we're waiting
for it to be paged in or out (typically to or from disk).

@@

Nor is a separate structure for each page desirable if there are
a large number of pages.  Since the state of one page is often
identical to the state of many others, a pagemap of pointers
suggests itself.

@@

\begin{verbatim}
struct pagemap_entry {
   /* ACCESSLIST
    * A list of clients that currently have access, along
    * with an indication if this is read-only or read-write access.
    */
   port_t clients_with_access[];
   bool write_access_granted;

@@

   /* INVALID
    * indicates that the backing store data is invalid because we got an
    * error return from a write attempt
    *
    * error returns on read or unlock operations do not set "invalid"
    */
   bool invalid;
};

Global structure {
   /* NEXTERROR list
    * A linked list of outstanding (error code, page, client) tuples.
    * They'll be transmitted to the client that requested them.
    */

@@

How to maintain the pagemap for a large number of clients?  Each page
has a pointer to a structure indicating which clients currently hold
copies of the page.  We anticipate that many pages (say, all the pages
in a single shared library) will exist in the same state, so there's
one structure for each combination of its entries, with multiple pages
pointing to it.  Pagemaps can be potentially large, so a pagemap entry
should be a single pointer.

Each file corresponds to a separate memory object, and separate memory
objects get separate control ports in the kernel.  A future
enhancement could be for the kernel to use a single control port for
all of its memory objects, allowing libpager to reduce the number of
these combination structures, if different files have the same usage
pattern (a program and some of its shared libraries, for example).

Each pagemap entry has an ACCESSLIST and a WAITLIST.  ACCESSLIST is
unordered; WAITLIST is ordered.  ACCESSLIST is stored sorted.

What happens when either list gets a new client (or loses one)?  We
have to search in the structures for the new combination, and create a
new structure if the desired combination doesn't exist yet.

We currently use a C++ std::set and let it do the searching for us.

\begin{comment}

Other possibilities:

  How can we organize the structures?  First, by the number of clients
  on each list.  We'll always know how many clients are in the
  ACCESSLIST we're searching for, so if we've got six ACCESSLIST clients
  and are moving the only client on WAITLIST to ACCESSLIST, we'll search
  in the seven-client ACCESSLIST, zero-client WAITLIST tables.  We can
  use a hash table or a tree.

  Because ACCESSLIST is sorted, we can binary search to find an
  ACCESSLIST that matches the one we want, then linear search forward
  and backward to find the one we want.

  or... Keep ACCESSLIST and WAITLIST completely separate for each other.
  Each pagemap entry has an ACCESSLIST pointer and a WAITLIST pointer,
  as well as some flags, including WRITE-ACCESS, which indicates if
  we've given out write access to a single client, in which case
  ACCESSLIST isn't a pointer, but a single client port number.  If
  WRITE-ACCESS is false, then ACCESSLIST points to a NULL-terminated
  list of client port numbers.  The disadvantage here is that the
  pagemap entries require more space (at least two pointers).

  WAITLISTs are ordered and include a flag for each client indicating
  whether it's waiting for READ access or WRITE access.  A special case,
  indicated by a flag, could be a single client on a WAITLIST, in which
  case we don't need a pointer to a WAITLIST, but rather can use the
  port number instead of the pointer.

The pagemap structures do not count the number of pages pointing to
them.  If we did that, it would facilitate deallocation when the count
drops to zero, but would interfere with the "add table" pointer
optimization described below.

POINTER OPTIMIZATION (UNIMPLEMENTED)

To speed this process, each structure can maintain pointers to
accelerate this mapping.  Each structure would have an "add table"
would maintain mappings like "add client 123 -> structure 0x80fadde0".
If the "add table" doesn't have an entry, then we need to search all
the structures to find one.  A problem with this is the need to
invalidate those pointers when a structure is deallocated.  Another
problem is that it requires the structures to resize when these
pointers are added. Keep this idea for a future enhancement.

single client pointer optimizations
  1. no access ->2 ->3
    pagemap_entry * put_client_1_on_waitlist_for_READ
    pagemap_entry * put_client_1_on_waitlist_for_WRITE
  2. waiting for read ->4
    pagemap_entry * move_first_WAITLIST_client_to_ACCESSLIST
  3. waiting for write ->5
    ditto (move_first_WAITLIST_client_to_ACCESSLIST)
  4. has read ->6
    ditto (put_client_1_on_WAITLIST_for_WRITE)
  5. has write
  6. has read; waiting for write ->5
    ditto (move_first_WAITLIST_client_to_ACCESSLIST)

PAGEMAP FOR A SMALL NUMBER OF CLIENTS (UNIMPLEMENTED)

I've thought of a possible optimization for small numbers of clients
that would largely avoid the need for separate pagemap structures, and
would transition to the unlimited size structures described above when
more clients connect that it can handle.

First, steal a bit from the pointer by aligning the structure on a
word boundary.  Then, if the least significant bit is 1, interpret the
pagemap entry as a bitfield rather than as a pointer to a more
complicated structure.

A 32 bit field (err... make that a 31 bit field) can handle up to seven
clients.  The new pagemap entries have an ACCESSLIST, which is an
unsorted set of clients (those who currently have access), and a
WAITLIST, which is a sorted list of clients that have requested
access, and a two boolean flags.

We number our clients 0 through 6, and I'm thinking above moving their
ports into high port number space to easily identify them.  So, the
first client on the first memory object would be on port 0x8000000,
the second client would be on 0x80000001, etc.  The second memory
object would have its ports on 0x80000008 through 0x8000000f.  So we
can just strip off the lowest 3 bits to figure out our "client
number".

We need a paging out flag, and a bit to indicate if the granted access
is read-only or read-write.  Assume that the {\tt error} field is
always {\tt KERN_SUCCESS}.  The ACCESSLIST requires one bit per client
to indicate if access is granted.  The WAITLIST is sorted (client
requests are processed FIFO) and requires four bits per slot -- three
to identify the client (all ones indicates an empty slot), and one to
indicate if the request is for read or write access.  Thus, we need
seven bits for the ACCESSLIST, two more flag bits (paging out and
read/write access), one bit to indicate that this isn't a pointer,
leaving room for five WAITLIST slots.

Once we hit our seventh client, or some oddball condition (like an
error return other than the expected KERN_SUCCESS, or a sixth client
on the WAITLIST), we shift to a pointer that points to a more
complicated, dynamically allocated structure.

Likewise, a 64-bit pointer (wishful thinking at the moment, since Hurd
is currently a 32-bit system) would allow fifteen clients, with an
eleven client WAITLIST.  Both the 32-bit and the 64-bit scheme have
two bits left over, perhaps they could be used to allow EIO, ENOSPC,
and EDQUOT, in addition to KERN_SUCCESS, since these seem to be the
most likely error codes.

\end{comment}

ACCESS CONTENTION

Putting a client on an empty queue triggers lock requests to all
clients in the working set, with the exception of the requesting
client itself, if it's already got read access and is requesting write
access.  Once the working set size drops to zero (or one), then the
original request can be satisfied.

In the meantime, additional clients can be added to the queue.  Each
client should be allowed access to a page for a period of time before
a lock request is sent.  Is the easiest way to do this to use the
timeout parameter on mach_msg?

Current implementation waits until the kernel returns a page, then
writes it to disk.  If a request for the page comes in before the
write completes, we wait for the write to finish, then send it right
back out without needing a read.

The new implementation will be handling pages as they're passed from
one client to another.  Should they be written out each time they are
handled?  Not write them at all until they're finally released?  Have
some kind of timer or counter to write them at a controlled pace?
Initially, we'll not write them at all until they're finally released,
which mimics the current behavior.

So when the last client returns a page, we look to see if anything is
in the wait queue.  If so, the first client(s) in the queue get the
pages.  If not, we flag that a write is in progress and start it.
When it completes, we check the queue again to see if anything is now
waiting.  If so, we service it.  If not, we discard the data and
notify the user that the page has been evicted.

ERROR HANDLING

@@

\section{Pseudocode}

@@

Old libpager would flush everything (no special flags set), then
destroy the receive right and ignore any final messages.  This created
a minor race condition where clients could request a page while the
flush was processing, which would then get dropped.

>>>>

\usepackage{nopageno}
\usepackage{vmargin}
\usepackage{comment}
\usepackage{xspace}
\usepackage{hyperref}
\usepackage{mdframed}
\usepackage{xcolor}

\definecolor{lightblue}{rgb}{0.8,0.8,1.0}

\usepackage{../slides/hurd}
%\usepackage{tikz}
\usepackage{tkz-graph}
\usetikzlibrary{positioning, fit, backgrounds, arrows, trees, shapes, arrows.meta}

@@

\title{Building an SSI Cluter with GNU/Hurd}
\author{Brent Baccala}
\date{May 2019}

@@

% Shortcuts just for this document
\def\libpager{{\tt libpager}\xspace}
\def\netmsg{{\tt netmsg}\xspace}
\def\pagersync{{\tt pager_sync()}\xspace}
\def\pagerwritepage{{\tt pager_write_page()}\xspace}
\def\pagerreadpage{{\tt pager_read_page()}\xspace}
\def\moinit{{\tt memory_object_init}\xspace}
\def\modatasupply{{\tt memory_object_data_supply}\xspace}
\def\modataerror{{\tt memory_object_data_error}\xspace}

@@

Operating system designers have tried for years to leverage the power
of multiple computers and develop cluster operating systems.  Notable
attempts include VAX/VMS, MOSIX, Beowulf, Plan 9, ScaleMP, and Hadoop.

@@

GNU/Hurd is an operating system based on the Mach kernel, which was
designed for use in a cluster environment.  Hurd is designed around a
set of servers that communicate primarily by remote procedure calls
mediated by the Mach kernel, making it a natural choice to build a
POSIX-compatible SSI cluster.  Development on Hurd has progressed
slowly, but a Debian port has been available since 2013, and
approximately three quarters of the Debian packages compile and run on
Hurd.

Hurd needs several enhancements to operate as an SSI cluster.  First,
Mach messages must be transported over network links, a function
provided by yet another Hurd server, called \netmsg.  Next,
distributed shared memory must be implemented, without which we can't
achieve full POSIX compatibility.  Fortunately, Mach's memory
management subsystem was designed to support distributed shared
memory, and Hurd's interface to Mach's memory management subsystem
is compartmentalized in a C library called \libpager.

\libpager is a crucial library.  It provides the cache coherence logic
necessary to implement inter-node shared memory, whether disk-based or
not.  It is used, for example, by the filesystem servers to implement
{\tt mmap()}.  Hurd's original \libpager, written in the mid-1990s,
only supported a single client (a client being a Mach kernel), and was
thus suitable for single node operation, but its API was designed to
eventually support multiple clients.  To support a multi-node Hurd
cluster, a new \libpager was written to support multiple clients (i.e,
multiple Mach kernels).  This allows both inter-node {\tt mmap()}
as well as inter-node POSIX shared memory.

Finally, Hurd's C library needs to be modified to allow cross-node
forks, and Hurd's authentication mechanism needs to be extended
as well.

Hurd's \netmsg server and the \libpager library have been developed
to the point where Hurd could be usable as a cluster operating system,
though the current Debian distribution doesn't use this functionality.
At the time of this writing (2019), the major barriers to building a
Hurd cluster are the 32-bit address space and the lack of SMP support,
enhancements that are beyond the scope of the work described in this
paper.

The CAP theorem states that a distributed system can't have all of
three things: consistency, availability, partitioning.  Right now, we
sacrifice availability -- if the network partitions, \libpager can
deadlock.  If we introduce a timeout on lock requests, then we're
sacrificing consistency.

@@

\begin{figure}
\begin{center}

\tikzset{
    state/.style={
        draw, ellipse
    },
    >=latex
}

\def\notm{$\overline{\rm m}$}
\begin{tikzpicture}[node distance=3cm]

\node (e) [state] {e};
\node (wait-r) [state, above left=of e] {wait-r};
\node (wait-w) [state, above right=of e] {wait-w};
\node (ww) [state, right=of wait-w] {\notm,w,w};
\node (rw) [state, above right=of wait-w] {\notm,r,w};
\node (nw) [state, below right=of wait-w] {\notm,n,w};
\node (nr) [state, left=of wait-r] {\notm,n,r};
\node (rr) [state, above left=of wait-r] {\notm,r,r};
\node (nn) [state, below left=of wait-r] {\notm,n,n};

\draw [->] (e) -- node[midway]{{\bf data_request}} node[midway, below]{(read)} (wait-r);
\draw [->] (e) -- node[midway]{{\bf data_request}} node[midway, below]{(write)} (wait-w);
\draw [->] (wait-r) -- (wait-w);
\draw [->] (wait-r) -- node[midway]{{\bf data_supply}} node[midway, below]{(no access)} (nr);
\draw [->] (wait-r) -- node[midway]{{\bf data_supply}} node[midway, below]{(read access)} (rr);
\draw [->] (wait-r) to[out=45, in=135] node[midway, above]{{\bf data_supply}} node[midway, below]{(write access)} (ww);

\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(no access)} (nw);
\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(read access)} (rw);
\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(write access)} (ww);

\draw [->] (wait-r) to[out=-135, in=-135] node[midway]{{\bf data_error}} (e);
\draw [->] (wait-w) to[out=-45, in=-45] node[midway]{{\bf data_error}} (e);

\end{tikzpicture}
\end{center}
\caption{Page-in transitions.  Adapted from figure 6 in [1]}
\label{Mach memory}
\end{figure}

Each virtual memory page in the Mach kernel can be in one of fifteen
different states: empty (e), waiting for either read or write access,
and twelve different states for pages in memory, depending on
whether the page has been modified or not, the current access, and the
desired access.  Figure \ref{Mach memory} shows the transitions
involved in paging in; [1] uses five similar diagrams to completely
illustrate all of the Mach memory manager's possible transitions.

Mach's ability to attach out-of-band memory to messages is thus a
crucial component in the efficient implementation of its memory
management.  A page is typically supplied to kernel by attaching it to
a \modatasupply message.  No copying needs be done, as the
page is transferred to kernel control simply by adjusting bits in the
hardware memory management tables.  Likewise, transferring a modified
page from the kernel to a user-space memory manager is done by
attaching it to a Mach message from the kernel.

@@

Each file corresponds to a separate memory object, and separate memory
objects get separate control ports in the kernel.  A future
enhancement could be for the kernel to use a single control port for
all of its memory objects, allowing \libpager to reduce the number of
these combination structures, if different files have the same usage
pattern (a program and some of its shared libraries, for example).

@@

\section{Introduction to Hurd}

The Free Software Foundation chose Mach in the early 1990s as the
kernel for what was envisioned as its flagship operating system, Hurd.
The name is an allusion to the collection of servers required to
implement the core functionality usually found in an operating system
kernel, but which are transferred to user space in the Mach design.
Linux ultimately eclipsed Hurd with a faster development process and a
simpler, more traditional design that packed functionality into a
large, monolithic kernel.  Hurd development has continued, however,
and it now available as a Debian port, though without the bevy of
features and device drivers that have been packed into Linux over the
years.

For example, the file system, usually implemented as part of the
kernel, is mostly decoupled in Hurd.  In principle, a server called
{\tt pci-arbiter} is responsible for mapping PCI devices, such as
a SCSI controller, into the memory space of user space programs that
manage them.  In reality, GNU Mach (the version of Mach used by Hurd)
contains a significant number of device drivers ripped from an older
version of the Linux kernel, so we do have some device drivers
(including our disk controllers) in the kernel.

Aside from reading and writing raw disk blocks, however, the file
system implementation is completely done in user-space servers, called
{\it translators} in Hurd nomenclature.  The most important file
server is {\tt ext2fs}, which interprets and manages an ext2 file
system.  The program interprets {\tt dir_lookup} RPCs and typically
responds to them by creating a port (in its own port space) that will
represent a file handle, then replies to the {\tt dir_lookup} message
with a send right to the port.  The file handle port handles messages
like {\tt io_stat}, {\tt io_read}, and {\tt io_write} to carry out
basic I/O, as well as messages like {\tt file_exec} (if the file is
executable) and {\tt file_chown} to change the file's ownership, and
some more exotic messages like {\tt io_async}, which requests that the
fileserver send messages to the process when I/O is available
(typically used for sockets rather than files), and {\tt
  io_reauthenticate}, which tells the file server to drop this port's
association with a user and perform an authentication exchange.

One of the advantages of this design is that mounting a loopback
device is a non-privileged operation.  All that is required is to run
another copy of {\tt ext2fs}, pass it the name of the file containing
the loopback filesystem, and attach it to a name in the directory
hierarchy.  Voila!  All future {\tt dir_lookup} calls into that part
of the filesystem will be relayed to the new {\tt ext2fs}.  Likewise,
an {\tt iso9660fs} translator is available that allows CD and DVD
images to be loopback mounted without special permissions.

A great deal of Hurd's POSIX compatible API is implemented in the
standard C library.  It is here, for example, that {\tt read()} is
implemented using an {\tt io_read} RPC exchange over a Mach port
linked to a Hurd server.  In addition to a program's standard threads
of execution, each program also has a thread, managed by code in the C
library, that listens on a message port.  POSIX signals, for example,
are delivered by sending message to this port.  Also, while the Linux
kernel can be queried to find a lot of information about a process,
such as the command string used to invoke it, the user and group
permissions it is running with, its environment, and its open file
descriptors, among other things, Hurd does not track any of this
information in the kernel.  In Hurd, much of this information is
obtained by querying the process directly, using its message port,
though this has the disadvantage of depending on the correct operation
of the program.  In particular, it is fairly easy to experience a
``hung'' {\tt ps} in Hurd, caused by a deadlocked program that is not
responding to its message port.

A sampling of some major Hurd servers include the following:

\begin{description}
\item[{\tt exec}]

is responsible for executing programs.  It
parses the ELF binary format and performs the necessary relocations
and memory mappings to prepare a new Mach task and make it ready to
run.  If the program begins with the hash-bang sequence ({\tt \#!}),
{\tt exec} finds a suitable interpreter to run the program.  Users can
specify their own {\tt exec} servers, so, for example, if a user
wanted to use {\tt qemu} to emulate a different architecture, no
kernel changes would be required, only a custom {\tt exec} server.

\item[{\tt auth}]

implements POSIX authentication based on users and groups.  A Hurd
process will be passed a send right to the {\tt auth} server when it
starts, and it is the possession of this send right that allows the
process access to protected resources.  The naive idea of simply
passing this send right to a server would grant the server all of the
process's access rights, so this is not done.  Instead, the process
creates a new port when it desires to authenticate to a server and
passes send rights to that port to both the server and the {\tt auth}
server.  The server then passes its copy of the send right to the {\tt
  auth} server, which matches the two against each other and confirms
that the desired access is permitted.  Matching the send rights is
trivial on a single node, since Mach guarantees that rights to the
same port will always appear on the same port number.

\item[{\tt proc}]

tracks processes in a POSIX-compatible way.  For
example, the Mach kernel maintains no sense of a parent/child
relationship between processes, or any concept of process ownership by
a user or group.  Killing a process (or sending any signal to a
process) requires an authentication step to obtain the process's
message port; this authentication is provided by the {\tt proc}
server.  Also, some information about a process (like its command name
and environment) needs to be provided without any authentication to
allow a program like {\tt ps} to operate; the {\tt proc} server
provides this functionality, as well.

\item[{\tt pfinet}]

is the Hurd's TCP/IP networking stack.  Networking functions in
the C library trigger RPC calls to this server.

\end{description}

\section{\netmsg}

Due to its heavy reliance on message passing between user-space
servers, Hurd is a natural choice to build an SSI cluster, if we can
transport Mach messages over a network.  Several programs have been
written over the years to do this; all collectively go under the
name of the ``netmsg server'', or just \netmsg.

I wrote Hurd's current \netmsg server in 2016.  It operates over a
single unsecured TCP/IP connection and has many limitations, but works
well enough to implement distributed shared virtual memory.  In server
mode, it presents a single node, by default the root of the
filesystem, to any clients that connect.  In client mode, it acts as a
translator, allowing a remote server's exported node to be mounted in
the local filesystem.

EXAMPLE: a \netmsg session

Thus, we have a remote file system, but it's important to realize that
this remote file system supports full Mach messaging.  Opening a file,
for example, is done by sending a Mach io_? message that is relayed
across the network, then received, processed, and answered by the
remote Hurd server managing that file.  This is in contrast to remote
file systems like NFS that don't transport Mach messages, and only
allow a subset of Hurd RPCs which can translated into NFS messages.
In particular, since Mach virtual memory is implemented using the Mach
messaging protocol, \netmsg allows a remote file to be {\tt mmap}'ed
into local memory.

\netmsg works by copying Mach messages almost verbatim across the
network connection.  Only the Mach port numbers are changed.
Out-of-band memory is simply copied across the network connection,
then deallocated.

While usable, this version of \netmsg has many flaws.  Transporting
all of the Mach messages over a single TCP/IP session provides
reliable, in-order delivery as guaranteed by Mach, but at the cost of
serializing all messages between two hosts.  Also, using the system's
TCP/IP stack requires context switches between \netmsg and the
TCP/IP server (a user-space program, like so much else in Hurd) for
every Mach message being relayed.  The current scheme doesn't do a
good job of handling more than two nodes, because there's no mechanism
to detect a cycle, which could result in a lot of unnecessary network
chatter.  Finally, it's become clear that the simplistic means of
managing ports creates unavoidable race conditions.

For example, consider (again) what happens when node A drops its last
send right to a port on node B.  Node A sends a NO SENDERS message to
node B and forgets about the port.  It's possible that node B has sent
node A another send right for the same port, and that it's in transit
when node A sends the NO SENDERS message.  In that case, node A would
actually have an outstanding send right when node B receives the NO
SENDERS message, which would cause node B to incorrectly drop the send
right linked to node A.  I can see no way around this except a more
complicated handshake where the NO SENDERS message from node A to node
B would include a sequence number allowing node B to determine that it
hadn't sent any additional send rights, and a reply message from node
B to node A finalizing the port deallocation.  (what about SEQNOS?)

\section{\libpager}

Hurd's implementation of the Mach memory management protocol is
implemented in a library called ``libpager''.  The original \libpager,
written in the mid-1990s, only supported a single client (i.e, a
single Mach kernel), and was thus suitable for single node operation.
Its API, however, is flexible enough to allow a more sophisticated,
multi-client \libpager as a drop-in replacement.  For dynamically
linked servers, all that is required is to update the system shared
library to obtain multi-node functionality.

@@

This routine is expected to allocate memory (unless it returns an
error) containing the requested data and pass it to \libpager, which
is responsible for deallocating it (typically by passing it to the
kernel).  A common use scenario is that {\tt pager_read_page()} reads
a page from disk and returns it.  {\tt pager_read_page()} can be
called from multiple threads to service multiple paging requests, and
must implement its own thread safety.  However, it is free to
serialize itself with a global mutex, and this will not prevent {\tt
  libpager} from servicing other requests.

@@

This routine is expected to save a modified memory page just before
\libpager releases its last copy of the data.  It does no memory
allocation or deallocation.  It receives allocated memory and leaves
it allocated, with \libpager responsible for dellocation, or passing
it to the kernel, if a new request has arrived in the time required
for {\tt pager_write_page()} to complete.

@@

How do we figure out which client is the local kernel?  One possible
solution is for \netmsg to listen on /servers/netmsg for an RPC that
accepts a send right and identifies which node currently hosts the
corresponding receive right.  Mach 3 Kernel Interfaces defined the
{\tt norma_port_location_hint} RPC that might be useful.

@@

DISTRIBUTED LIBPAGER.  We could use a shim process on remote nodes, to
allow requests for an existing page go to a node that holds a copy of
that page, instead of to the node with the disk.  Two ways I can
imagine this.  Either ext2fs forks off multiple processes on different
nodes, or a mechanism is developed to detect when a port is remote,
figure out which node it (currently) resides on, and then \libpager can
fork a process and push it to the other node, invisible to ext2fs
proper.

\subsection{Configuration Options}

Several \libpager implementation choices exist, and perhaps should be configuration options.

@@

  Another possible future enhancement is to track how many calls
  to \pagerreadpage are outstanding, and transition away from calling
  this routine if it appears to be overloaded.

@@

\item Are the pager_ functions thread-safe?  Do we call them from
  another thread before they return?

  {\tt pager_write_page()} is called in a protected manner -- only
  from a single thread (but not always the same thread), and \libpager
  waits until it returns before calling it again.

  Both {\tt pager_read_page()} and {\tt pager_unlock_page()} may be
  called reentrantly while another thread is waiting for another call
  to finish, but never on the same page.

@@

\subsection{The Pagemap}

All Mach memory management operations operate on memory pages, so
\libpager needs to maintain a table that tracks the state of each
page, called the pagemap.  Hurd's original \libpager had a relatively
simple pagemap.  The pagemap was an array of 16-bit shorts; each page
had a single 16-bit entry, allocated as follows:

\begin{figure}[h]
\begin{mdframed}[backgroundcolor=lightblue]
\begin{verbatim}
#define PM_WRITEWAIT  0x0200   /* queue wakeup once write is done */
#define PM_INIT       0x0100   /* data has been written */
#define PM_INCORE     0x0080   /* kernel might have a copy */
#define PM_PAGINGOUT  0x0040   /* being written to disk */
#define PM_PAGEINWAIT 0x0020   /* provide data back when write done */
#define PM_INVALID    0x0010   /* data on disk is irrevocably wrong */

@@

enum page_errors
{
  PAGE_NOERR,
  PAGE_ENOSPC,
  PAGE_EIO,
  PAGE_EDQUOT,
};
\end{verbatim}
\end{mdframed}
\caption{Pagemap structure in {\it original} {\tt libpager}}
\end{figure}

Two bits were allocated to each of two error codes.  {\tt ENOSPC} and
{\tt EDQUOT} were preserved; all other errors get collapsed into {\tt
  EIO}.  The remaining bits indicate if the page is in core, if we're
waiting for it to be paged in or out to the backing store.

@@

Nor is a separate structure for each page desirable if there are a
large number of pages.  We anticipate that many pages (say, all the
pages in a single shared library) will exist in the same state, so
there's one structure for each combination of its entries, with
multiple pages pointing to it.  Pagemaps can be quite large, so a
pagemap of pointers suggests itself.

@@

\begin{figure}[h]
\begin{mdframed}[backgroundcolor=lightblue]
\begin{verbatim}
struct pagemap_entry {
   /* ACCESSLIST
    * A list of clients that currently have access, along
    * with an indication if this is read-only or read-write access.
    */
   port_t clients_with_access[];
   bool write_access_granted;

@@

   /* INVALID
    * indicates that the backing store data is invalid because we got an
    * error return from a write attempt
    *
    * error returns on read or unlock operations do not set "invalid"
    */
   bool invalid;
};
\end{verbatim}
\end{mdframed}
\caption{The new pagemap structure}
\end{figure}

\begin{verbatim}
Global structure {
   /* NEXTERROR list
    * A linked list of outstanding (error code, page, client) tuples.
    * They'll be transmitted to the client that requested them.
    */

@@

\subsection{Access Contention}

The primary data structures for mediating contention between clients
are the ACCESSLIST and the WAITLIST in the pagemap.

Putting a client on an empty WAITLIST triggers lock requests to all
clients on the ACCESSLIST, with the exception of the requesting client
itself, if it's already got read access and is requesting write
access.  Once the ACCESSLIST drops to zero (or one, in the exceptional
case just described), then the original request can be satisfied.

In the meantime, additional clients can be added to the WAITLIST,
which basically triggers nothing.  The rule is that if the WAITLIST is
not empty, then we're already trying to revoke access to the clients
on ACCESSLIST.  Remember, there's currently no guarantee that a client
will have access for any period of time at all.

When a client with write access returns a page, we look to see if
anything is on the WAITLIST.  If so, the first client(s) on WAITLIST
get the pages.  If not, we flag that a write is in progress and start
it.  When it completes, we check the WAITLIST again to see if anything
is now waiting.  If so, we service it.  If not, we discard the data
and notify the user that the page has been evicted.

If multiple clients with READ access attempt to unlock near
simultaneously, will the data_unlock logic will trigger multiple lock
requests (a set for each client attempting to unlock)?

\subsection{Error Handling}

@@

Handling error on unlock requests is somewhat tricky.  {\it Mach 3
  Kernel Interfaces} states, in its {\bf NOTES} on {\bf
  memory_object_lock_request}:

\begin{quote}
When a running thread requires an access that is currently prohibited,
the kernel issues a {\bf memory_object_data_unlock} call specifying
the access required. The memory manager can then use {\bf
  memory_object_lock_request} to relax its access restrictions on the
data.

To indicate that an unlock request is invalid (that is, requires
permission that can never be granted), the memory manager must first
flush the page. When the kernel requests the data again with the
higher permission, the memory manager can indicate the error by
responding with a call to {\bf memory_object_data_error}.
\end{quote}

Thus, when {\tt pager_unlock_page()} returns an error, we can't simply
relay that error code in a reply to the kernel.  Instead, we must
flush the page and wait for a {\bf memory_object_data_request} from
that client, at which point we return the error code from {\tt
  pager_unlock_page()} rather than processing the data request
normally.  This is the function of the NEXTERROR list.  In the
original \libpager design, it was encoded in the pagemap with two
bits, since there was only one client, and all errors except ENOSPC
and EDQUOT were collapsed into EIO.  In the new design, NEXTERROR is
maintained completely separate from the pagemap, largely due to the
expectation that it will seldom be used.

For example, consider what happens if client A has read access to
page, and is thus alone on ACCESSLIST.  Client B now requests read
access to the same page, so it goes on WAITLIST and we
call \pagerreadpage.  Before this call completes, client A requests
write access via an unlock request, so it goes on the WAITLIST.
Finally, \pagerreadpage returns an error.  What do we do?
We send an \modataerror to client B, and a lock request to
client A, with a queued NEXTERROR for client A.

\subsection{Pseudocode}

@@

Old \libpager would flush everything (no special flags set), then
destroy the receive right and ignore any final messages.  This created
a minor race condition where clients could request a page while the
flush was processing, which would then get dropped.

\subsection{Flushing Pages}

Mach kernels are allowed to proactively flush pages unless the
``precious'' flag is specified, so we can't be certain that a client
on a page's ACCESSLIST actually possesses that page.  All we know for
sure is that a client not on ACCESSLIST does not possess the page.
